---
title: "Data Visualizaton Project Write-Up"
output:
  pdf_document: default
  html_document: default
date: "2024-04-03"
---

# Importing neccesary packages

```{r}
suppressMessages({
  library(tidyverse)
  library(ggcorrplot)
  library(dplyr)
  library(corrplot)
  library(ggplot2)
  library(gridExtra)
  library(cluster)
})
```


# Introduction

Information about the dataset can be found at: https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-10-17/readme.md

To simplify unnecessarily complexity, we propose excluding these columns. We believe they are not particularly interesting or relevant to how people score the albums. Our focus is on studying the statistics of Taylor Swift's songs based on audio features and how these features influence album scoring by listeners.

| Column Name         | Data Type | Description                   |
|---------------------|-----------|-------------------------------|
| ep                  | logical   | Is it an EP?                 |
| album_release       | double    | Album release date           |
| artist              | character | Artists                       |
| featuring           | character | Artists featured              |
| bonus_track         | logical   | Is it a bonus track?         |
| promotional_release | double    | Date of promotional release  |
| single_release      | double    | Date of single release       |
| track_release       | double    | Date of track release        |

The overall dataset is as follows:

- Three columns represent the index

- 11 numerical features

- 6 categorical features

- 2 target numerical variables

According to our proposal, we would like to conduct a series of necessary data mining and visualization tasks to extract meaningful insights from this dataset. Considering the characteristics of the dataset, we believe it is reasonable to perform the following analytics:

1. Descriptive Statistics (for both numerical and categorical features)

2. Features Distribution by scores

3. Time-series Analysis (based on the album release date)

4. Correlation Analysis (for the numerical features)

5. Regression Analysis 

With the hypothesis that audio features (numerical and categorical values) and the scores for each album have an inferential relationship, we will conduct our Regression Analysis using the audio features as input variables and scores as output variables. We will focus on two aspects:

- Since we lack scores for individual songs, it is reasonable to assign each song the score from "taylor_album_songs.csv" based on the score for the entire album in "taylor_albums.csv."

- One might suggest that we can aggregate all song features to represent an album, using that representation as input features to predict the scores. For simplicity, we will use the mean and median as the statistical measures to aggregate all the songs. However, the current size of the dataset that we can retrieve is relatively small (~12 valid records), which theoretically can be challenging to fit a regression model. Hence, we would like to discuss the potential of using such representation without fitting a specific model in our implementation.

# Answering the questions
## Question 1: How are the audio features distributed?
### Approach

To understand the distribution of audio features, our approach is structured into three main steps. 
1. **Descriptive Statistics** (Section 1): Initially, we delve into the distribution of the audio features through descriptive statistics to get a broad overview of their general characteristics.
2. **Feature Distribution by Scores** (Section 2): Subsequently, we explore how the distribution of features varies with scores, paying particular attention to the distribution of categorical features with respect to scores.
3. **Time-Series Analysis of Scores** (Section 3): Lastly, we analyze the scores over time to identify any temporal patterns or trends in the data.

Our analysis reveals that Section 2 contains some of the most compelling insights, particularly in relation to how categorical features are distributed across different scores.

Given the presence of multiple categorical features, and our interest in examining how scores are distributed for each category, we find that facet grids are particularly well-suited for this analysis. Facet grids offer a concise and comprehensive graphical representation that encompasses all categories, making it easier to identify patterns and outliers.

### Analysis
Please refer to Section 2 for additional details on the code, graphs, and findings.


### Discussion

The graph clearly illustrates that songs in a major mode tend to have higher scores than those in a minor mode, indicating a preference for Taylor’s compositions in a major key. This finding suggests a pattern in the composition of well-received music, particularly in the context of the artist in question. Furthermore, it is observed that, as Taylor roots in country music, the most commonly used keys are G, C, E, D, and A. This choice of keys may reflect both the genre's traditional sound and her personal style and preferences.


## Question 2: How do numerical audio features of each song affect the overall rating?
### Approach
To examine the relationship between audio features and song ratings, we employ a correlation matrix (Section 4). This matrix helps in identifying how each audio feature correlates with the ratings, providing insights into which features have the most significant impact on the overall score.

The correlation between two variables is calculated using the Pearson correlation coefficient, which is represented by the formula:

$$ r = \frac{n(\sum xy) - (\sum x)(\sum y)}{\sqrt{[n\sum x^2 - (\sum x)^2][n\sum y^2 - (\sum y)^2]}} $$

Where:

- `r` is the correlation coefficient,

- `n` is the number of observations,

- `x` and `y` are the variables being analyzed,

- $\sum xy$ is the sum of the product of paired scores,

- $\sum x$ and $\sum y$ are the sums of the scores, and

- $\sum x^2$ and $\sum y^2$ are the sums of the squared scores.

### Analysis
Please refer to Section 4 for additional details on the code, graphs, and findings.

### Discussion

The analysis of the correlation matrix reveals the following features to be most correlated with the `metacritic_score`: `user_score`, `loudness`, `acousticness`, `energy`, `duration_ms`, and `liveness`.

Moreover, when focusing on the `user_score`, the features with the highest correlation include `metacritic_score`, `speechiness`, `duration_ms`, `danceability`, `acousticness`, and `liveness`.

It becomes evident from this analysis that attributes such as `danceability`, `acousticness`, and `liveness` play significant roles in how users rate songs. These features, indicative of the song's rhythm, natural sound, and the presence of a live audience, respectively, are crucial in determining a song's appeal to listeners. This insight underscores the importance of these audio characteristics in the composition and production of music aimed at achieving high user ratings.


# Dataset Preprocessing
Firstly, we need to read the two csv files. The first file contain the information about audio features (numerical and categorical) for each song whereas the second file contain the information about the album name, release date and 
```{r}
album_song_df = read.csv("taylor_album_songs.csv")
album_df = read.csv("taylor_albums.csv")
```

Dropping unnecessary columns: 
```{r}
album_df_subset <- album_df[c("album_name", "metacritic_score", "user_score")]
album_song_df_subset <- album_song_df[c("album_name", "track_number", 
                                        "track_name", "danceability", 
                                        "energy", "key", "loudness", 
                                        "mode", "speechiness", 
                                        "acousticness", "instrumentalness", 
                                        "liveness", "valence", "tempo", 
                                        "time_signature", "duration_ms", 
                                        "explicit", "key_name", 
                                        "mode_name", "key_mode")]
```


We further remove rows with NaN values since we believe it is best to have a clean dataset. We don't think there is a reasonable way to interpolate these missing values, especially since they are subjective (for example, the user scores and Metacritic scores).

```{r}
album_df <- na.omit(album_df)
```


Assuming that metacritic score and user score are calculated as their definition, which are averaged over the songs and acts as a mean, we can use it to represent the score for each song in the same album. We remove all NaN rows since there is no reasonable way to interpolate these scores based on the other albums. And the number of NaN rows is insignificant.

```{r}
album_song_with_scores_df <- merge(album_song_df_subset, album_df_subset, 
                                   by = "album_name", all.x = TRUE)

album_song_with_scores_df <- na.omit(album_song_with_scores_df)

```
The constructed dataframe is now as follows:
- Three columns represent the index
- 11 numerical features
- 6 categorical features
- 2 target numerical variables

```{r}
# 3 first index columns

# 11 numerical columns
numerical_feats = c("danceability", "energy", "loudness", "speechiness", 
                    "acousticness", "instrumentalness", "liveness", 
                    "valence", "tempo", "time_signature", "duration_ms")

# 6 categorical columns
categorical_feats = c("explicit", "key", "key_name", "mode",
                      "mode_name", "key_mode")

# 2 target numerical columns: metacritic_score VS user_score
```

# 1. Descriptive Statistics

## 1.1. Categorical input variables

```{r}
categorical_feats_df <- album_song_with_scores_df[categorical_feats]

categorical_feats_df <- na.omit(categorical_feats_df)


# Convert categorical variables to factors
categorical_feats_df <- lapply(categorical_feats_df, factor)

# Get frequency tables for each categorical variable
frequency_tables <- lapply(categorical_feats_df, table)

# Plot bar charts for each categorical variable
for (i in seq_along(frequency_tables)) {
  # Set up plotting area
  par(mar = c(5, 5, 6, 2))  # Adjust margins for x-axis label
  
  # Plot bar chart with wider spacing between categories and 
  # horizontal x-axis labels
  barplot(frequency_tables[[i]], 
          main = paste("Frequency of", names(frequency_tables)[i]),
          ylab = "Frequency",
          col = "skyblue", 
          border = "black",
          space = 0.5,  # Adjust the spacing between bars
          las = ifelse(names(frequency_tables)[i] == "key_mode", 2, 1))  
  # Rotate x-axis labels vertically if it's "key_mode"
}

```

## 1.2. Numerical input variables
```{r}
numerical_feats_df <- album_song_with_scores_df[numerical_feats]

numerical_feats_df <- na.omit(numerical_feats_df)

summary(numerical_feats_df)
```

## 1.3. Target variables
### 1.3.1. The descriptive STAT for the scores per album
```{r}
target_df <- album_df[c("metacritic_score", "user_score")]

target_df <- na.omit(target_df)

summary(target_df)
```
### 1.3.2. Scores for each album 

```{r}
# Convert album_release to Date format
album_df$album_release <- as.Date(album_df$album_release)

# Plot Metacritic scores
ggplot(album_df, aes(x = album_name, y = metacritic_score)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Metacritic Scores of Albums",
       x = "Album Name",
       y = "Metacritic Score")

# Plot User scores
ggplot(album_df, aes(x = album_name, y = user_score)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "User Scores of Albums",
       x = "Album Name",
       y = "User Score")
```
It seems that there is a strong correlation between the two scores. The patterns look almost the same, with the user scores exhibiting more consistent distribution.

```{r}
# Plot box plot for Metacritic scores
ggplot(album_df, aes(x = "Metacritic Score", y = metacritic_score)) +
  geom_boxplot() +
  labs(title = "Box Plot of Metacritic Scores",
       x = "",
       y = "Metacritic Score") +
  theme_minimal()

# Plot box plot for User scores
ggplot(album_df, aes(x = "User Score", y = user_score)) +
  geom_boxplot() +
  labs(title = "Box Plot of User Scores",
       x = "",
       y = "User Score") +
  theme_minimal()
```

# 2. Feature distribution by scores

We employ a facet grids to visualize the distribution of the categorical features by scores.

```{r fig.height=8}
# Define the list of categorical columns
categorical_cols <- c("key", "mode", "explicit", "key_name", "mode_name")

# Create facet grid plots for each score
for (score in c("metacritic_score", "user_score")) {
  # Initialize a list to store individual plots
  plots <- list()
  
  # Loop through each categorical column
  for (col in categorical_cols) {
    # Convert the column to factor with explicit levels
    album_song_with_scores_df[[col]] <- 
      factor(album_song_with_scores_df[[col]],
            levels = unique(album_song_with_scores_df[[col]]))
    
    # Create a facet grid plot for the current categorical column
    p <- ggplot(album_song_with_scores_df, aes_string(x = col, y = score)) +
      geom_boxplot() +
      labs(x = col, y = score) +
      ggtitle(paste("Distribution of", score, "by", col))
    
    # Add the plot to the list
    plots[[col]] <- p
  }
  
  # Combine plots into a single facet grid
  grid.arrange(grobs = plots, nrow = 3, ncol = 2)  
  # Adjust nrow and ncol as needed
}

```

Some findings from the metacritic_score and the user_score graphs:

- Songs with explicit lyrics and a major mode seem to have higher scores than implicit ones.

- Songs in the keys of G, E, and F# have the highest scores.

- The most common keys used in Taylor's songs are G, C, E, D, and A, which reflect her style of using country music.

# 3. Time-series analysis

## Album scores over time

```{r}
album_df <- na.omit(album_df)

# Convert album_release to Date format
album_df$album_release <- as.Date(album_df$album_release, format = "%Y-%m-%d")

# Convert album_release to character for the x-axis
album_df$album_release_char <- as.character(album_df$album_release)

# Plotting Metacritic scores
ggplot(album_df, aes(x = album_release_char, y = metacritic_score)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Metacritic Scores of Taylor's Albums Over Time",
       x = "Album Release",
       y = "Metacritic Score") +
  scale_x_discrete(labels = album_df$album_release_char) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Plotting User scores
ggplot(album_df, aes(x = album_release_char, y = user_score)) +
  geom_col(fill = "darkorange") +
  theme_minimal() +
  labs(title = "User Scores of Taylor's Albums Over Time",
       x = "Album Release",
       y = "User Score") +
  scale_x_discrete(labels = album_df$album_release_char) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

# 4. Correlation analysis
## 4.1. Correlation map
```{r}
numerical_feats_and_scores_df <- album_song_with_scores_df[c("danceability", 
                                                             "energy", 
                                                             "loudness", 
                                                             "speechiness", 
                                                             "acousticness", 
                                                             "instrumentalness", 
                                                             "liveness", 
                                                             "valence", 
                                                             "tempo",
                                                             "time_signature", 
                                                             "duration_ms",
                                                             "metacritic_score",
                                                             "user_score")]

numerical_feats_and_scores_df <- na.omit(numerical_feats_and_scores_df)

# Compute the correlation matrix
correlation_matrix <- cor(numerical_feats_and_scores_df)

# Visualize the correlation matrix using corrplot
corrplot(correlation_matrix, method = "circle")
```
## 4.2. Comments

We can see that the most correlated columns w.r.t the target columns are:

```{r}
# Filter out the most correlated columns with respect to "metacritic_score"
metacritic_correlation <- abs(correlation_matrix["metacritic_score", ])
most_correlated_metacritic <- names(sort(metacritic_correlation,
                                         decreasing = TRUE)[2:7]) 
# Excluding self-correlation

# Filter out the most correlated columns with respect to "user_score"
user_correlation <- abs(correlation_matrix["user_score", ])
most_correlated_user <- names(sort(user_correlation, decreasing = TRUE)[2:7]) 
# Excluding self-correlation

# Display the most correlated columns
cat("Most correlated columns with respect to metacritic_score:\n")
cat(most_correlated_metacritic, "\n")

cat("Most correlated columns with respect to user_score:\n")
cat(most_correlated_user, "\n")
```
```{r}
# Select most correlated columns with 
# respect to "metacritic_score" excluding "user_score"
selected_columns_ms <- setdiff(most_correlated_metacritic, 
                                       "user_score")
selected_columns_ms <- c(selected_columns_ms, 
                                 "metacritic_score")

# Create subset dataframe with selected columns for "metacritic_score"
selected_numerical_feats_and_ms_df <- album_song_with_scores_df[selected_columns_ms]

# Select most correlated columns with respect to "user_score" 
# excluding "metacritic_score"
selected_columns_user <- setdiff(most_correlated_user, 
                                 "metacritic_score")
selected_columns_user <- c(selected_columns_user, "user_score")

# Create subset dataframe with selected columns for "user_score"
selected_numerical_feats_and_user_score_df <- 
  album_song_with_scores_df[selected_columns_user]
```



# 5. Regression Analysis 

## 5.1. Regression Analysis using numerical values

In this section, we proceed with the columns that have the highest correlation with the scores to use as input features and utilize metacritic_score and user_score as our predicted variables. 

We will utilize scatter plots to visualize the data points for each input variable and the target scores.

### 5.1.1. Regression analysis on metacritic_score

```{r}
# Perform linear regression analysis
regression_model <- lm(metacritic_score ~ ., data = selected_numerical_feats_and_ms_df)

# Summarize the regression results
summary(regression_model)
```
```{r}

# Create a scatter plot of metacritic_score against each predictor variable
plot_data <- selected_numerical_feats_and_ms_df
predictor_variables <- setdiff(names(plot_data), "metacritic_score")

for (var in predictor_variables) {
  # Create scatter plot for each predictor variable
  plot <- ggplot(plot_data, aes_string(x = var, y = "metacritic_score")) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = paste("Scatter Plot of", var, "vs. Metacritic Score"),
         x = var, y = "Metacritic Score")
  
  # Display each plot individually
  print(plot)
}
```


### 5.1.2. Regression analysis on user_score
```{r}
# Perform linear regression analysis
regression_model <- lm(user_score ~ ., data = 
                         selected_numerical_feats_and_user_score_df)

# Summarize the regression results
summary(regression_model)
```


```{r}

# Create a scatter plot of metacritic_score against each predictor variable
plot_data <- selected_numerical_feats_and_user_score_df
predictor_variables <- setdiff(names(plot_data), "user_score")

for (var in predictor_variables) {
  # Create scatter plot for each predictor variable
  plot <- ggplot(plot_data, aes_string(x = var, y = "user_score")) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = paste("Scatter Plot of", var, "vs. User Score"),
         x = var, y = "User Score")
  
  # Display each plot individually
  print(plot)
}
```
### 5.1.3 Comments

- The standard error is lower than that of the model predicting the metacritic score.

- There appears to be no linear relationship between the input features and user scores. It is evident that a more complex fitting method should be employed instead of a simple linear regression model.

- Acousticness and danceability can serve as decisive features in certain edge cases. When the acousticness exceeds 0.5, there is a high probability of assigning high user scores. Danceability associated with high user scores typically falls between 0.4 and 0.7.

# 6. Clustering

```{r}
# Select numerical features for clustering
features <- selected_numerical_feats_and_user_score_df[, c("danceability", 
                                                           "acousticness")]

# Perform K-means clustering
kmeans_result <- kmeans(features, centers = 3, nstart = 30)

# Add cluster labels to the data
selected_numerical_feats_and_user_score_df$cluster <- 
  as.factor(kmeans_result$cluster)

# Plot the clustered data
ggplot(selected_numerical_feats_and_user_score_df, aes(x = danceability, 
                                                       y = acousticness, 
                                                       color = cluster)) +
  geom_point() +
  geom_point(data = as.data.frame(kmeans_result$centers), 
             aes(x = danceability, y = acousticness), color = "black",
             size = 3, shape = 17) +
  labs(title = "K-means Clustering Result")

# Print cluster centers
print(kmeans_result$centers)
```


